---
title: "RDD Hansen Replication"
author: "Hannah Jones"
date: "2/15/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
library(haven)
library(tidyverse)
library(stargazer)
library(estimatr)
library(rdd)
library(rdrobust)
library(rddensity)
library(ggplot2)

dwi <- read.csv('https://raw.githubusercontent.com/hannahjonesut/causal-inference-class/master/Data/hansen_dwi.csv')
```

## Summary

The paper is studying the effect of severity of punishment on recidivism.  In this case, the author is looking at drunk driving and the impact of punishments that increase in severity with the increase in blood-alcohol level.  The paper aims to provide some basis for analyzing whether harsher BAC cutoffs and punishments will dis-incentivize the commitment of the offense (drunk driving).  The author is using data from ~500k DUI stops from the state of Washington from 1995 to 2011.  Notably, Washington instituted a 0.08 cut-off for DUI and 0.15 for aggravated DUI in 1999.  So, the author uses the 1999-2007 to look at the impact of having a BAC of 0.08 or 0.15 on recidivism.  The author is looking at discrete BAC thresholds of 0.08 and 0.15 to determine the impact of punishment on recidivism, as a BAC of 0.08 will result in punishment, and a BAD of 0.15 will result in harsher punishment.  The author finds that punishment serves as a good deterrent for drunk driving.  This suggests that lowering the BAC cutoff from 0.08 to 0.05 will be effective in deterring drunk driving behavior and recidivism.

## Replication

3. 
```{r}

#create a dummy variable for BAC >= 0.08
dwi <- dwi %>%
  mutate( 
    bac_high  = if_else(bac1 >= 0.08, 1, 0))
```

4. 
```{r}
#sort on running variable to see if any manipulation
d1 = dwi %>%
  group_by(bac1)%>%
  summarize(sum_recid = sum(recidivism))

ggplot(data = d1) +
  geom_col(aes(x = bac1, y = sum_recid)) +
  labs(
    title = "Frequency of Recidivism vs Blood Alcohol Content Histogram",
    caption = "Based on administrative records from the Washington State Impaired Driver Testing Program, 1999–2007.Vertical line at BAC = 0.08.",
    x="Blood Alcohol Content",
    y="Frequency") +
  geom_vline(xintercept = .08, colour = "red", linetype = 1) 
```

5. 
